---
title: "Image Compression using Principal Component Analysis"
author: "Nishit Jain"
date: "Built on: `r format(Sys.Date(), '%m/%d/%Y')`"
version: "v1"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 4
  html_notebook:
    toc: true
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# Load packages
required_packages <- c("tidyverse",
                       "jpeg",
                       "imager",
                       "raster",
                       "factoextra")

new_packages <- required_packages[!(required_packages %in% installed.packages()[, "Package"])]

# Installing the missing packages
if (length(new_packages))
  install.packages(new_packages, repos='http://cran.us.r-project.org')

# Load all required packages
lapply(required_packages, library, character.only=T)

options(scipen=999)

source("./compression_utils.R")
```

# Objective

The objective of this notebook is:

* To document research and implementation of Principal Component Analysis for image compression

* To demonstrate image compression using PCA

# Research

Principal Component Analysis is a way to identify and express patterns in data so as to highlight similarities and differences within it.

**This is extremely important for higher dimensional data which can not be visualized to identify patterns.**

After the pattern is identified, it can be used to compress the data to a lower dimension without much loss of information. This property of PCA is used for **image compression**.

PCA is based on eigendecomposition of covariance matrix of the data.

## Covariance

Covariance is a measure of the extent to which corresponding elements from two sets of ordered data move in the same direction.

Let $X$ and $Y$ be two random variables, then covariance is computed as:

$$
Cov(X, Y) = \frac{\sum_n{E[(X - E[X])(Y - E[Y])]}}{n-1}
$$

where $n$ is the number of items in the set

## Covariance Matrix

Covariance values for a set of variables are displayed as a covariance matrix, $C$, where

$$
C = \begin{bmatrix}
c_{1, 1} & c_{1, 2} & c_{1, 3} & ... & c_{1, n}\\
c_{2, 1} & c_{2, 2} & c_{2, 3} & ... & c_{2, n}\\
c_{3, 1} & c_{3, 2} & c_{3, 3} & ... & c_{3, n}\\
\vdots & \vdots & \vdots & ... & \vdots\\
c_{n, 1} & c_{n, 2} & c_{n, 3} & ... & c_{n, n}\\
\end{bmatrix}
$$

where $c_{i, j}$ is given as

$$
c_{i, j} = Cov(X_{i}, X_{j})
$$

**Note:** The diagonal elements, $c_{i, i}$ give $Var(X_{i})$

**Note:** Covariance matrix is symmetric as $c_{i, j} = c_{j, i}$

## Eigendecomposition

In linear algebra, eigendecomposition is the factorization of a matrix into a canonical form, whereby the matrix is represented in terms of its eigenvalues and eigenvectors.

A vector $v$ of dimension $N$ is an **eigenvector** of a square $N × N$ matrix $A$ if it satisfies the linear equation

$$
A\vec{v} = \lambda \vec{v}
$$
where $\lambda$ is any scalar value and corresponds to the **eigenvalue** for the respective **eigenvector**

**Note:** $\vec{v}$ needs to be a non-zero vector

Geometrically, this equation implies that for any matrix $A$, $\vec{v}$ represents any vector which, when linearly transformed using $A$, retains its direction but is scaled by a factor of $\lambda$.

## PCA Approach

PCA is performed using the following steps:

* Step 1: Represent the data as a $M X N$ matrix, where $M$ are the number of samples and $N$ are the number of features (dimensions)

* Step 2: Center the data by subtracting the mean of each feature across samples from the value of that feature for each sample

* Step 3: Calculate covariance matrix for the centered data

* Step 4: Perform eigendecomposition on the covariance matrix

* Step 5: Arrange the eigenvectors and eigenvalues in decreasing order of eigenvalues

* Step 6: Select top $l$ eigenvectors based on eigenvalues where $l < N$

## Image Compression

After selecting top $l$ eigenvectors based on eigenvalues of the covariance matrix of data, they can be used to compress images as follows:

* Step 1: Perform matrix multiplication of original data with selected eigenvectors to get lower dimension representation of data

* Step 2: Reconstruct original data using matrix multiplication of lower dimension data with inverse of matrix of eigenvectors

The reconstructed data will be a compressed version of original data

# Analysis

In this section we will demonstrate PCA based image compression on a real image.

## Load Image

We are using `jpeg::readJPEG()` function to load image from local directory.

```{r}
file_path <- "./sample_images/nishit_jain.jpeg"
loaded_image <- jpeg::readJPEG(file_path, native = FALSE)
plotImage(list(loaded_image))
```

## Image Properties

We are looking at the size and the dimensions of the original image.

```{r, echo=FALSE, results='asis'}
cat("* ", "Size of original image is ", file.info(file_path)$size, " bytes.\n\n", sep='')
cat("* ", "Dimensions of the image are: ", dim(loaded_image)[1], " X ", dim(loaded_image)[2], " X ", dim(loaded_image)[3], "\n\n", sep='')
cat("* ", "This image can be thought of as having ", dim(loaded_image)[1], " samples of ", dim(loaded_image)[2], " dimensions each with ", dim(loaded_image)[3], " different compression tasks.\n\n", sep='')
```

## Image Channels

```{r, echo=FALSE}
loaded_image_copy <- as.array(loaded_image)

loaded_image_copy[, , 2] <- matrix(0L, nrow=dim(loaded_image_copy[, , 2])[1], ncol=dim(loaded_image_copy[, , 2])[2])
loaded_image_copy[, , 3] <- matrix(0L, nrow=dim(loaded_image_copy[, , 3])[1], ncol=dim(loaded_image_copy[, , 3])[2])

image_r <- loaded_image_copy

loaded_image_copy <- as.array(loaded_image)

loaded_image_copy[, , 1] <- matrix(0L, nrow=dim(loaded_image_copy[, , 1])[1], ncol=dim(loaded_image_copy[, , 1])[2])
loaded_image_copy[, , 3] <- matrix(0L, nrow=dim(loaded_image_copy[, , 3])[1], ncol=dim(loaded_image_copy[, , 3])[2])

image_g <- loaded_image_copy

loaded_image_copy <- as.array(loaded_image)

loaded_image_copy[, , 1] <- matrix(0L, nrow=dim(loaded_image_copy[, , 1])[1], ncol=dim(loaded_image_copy[, , 1])[2])
loaded_image_copy[, , 2] <- matrix(0L, nrow=dim(loaded_image_copy[, , 2])[1], ncol=dim(loaded_image_copy[, , 2])[2])

image_b <- loaded_image_copy

plotImage(list(image_r, image_g, image_b))
```

Since we have 3 channels (R, G, B), we will perform PCA for each channel individually.

## Principal Component Analysis

In this section, we will use PCA to extract new dimensions from the image and plot their importance.

```{r}
# Extract 3 channels
loaded_image_r <- loaded_image[, , 1]
loaded_image_g <- loaded_image[, , 2]
loaded_image_b <- loaded_image[, , 3]
```

While performing PCA, we will not center the value for each pixel. This will enable us to reconstruct the original image after compression.

```{r}
# PCA on each channel
loaded_image_r_pca <- stats::prcomp(loaded_image_r, center=FALSE)
loaded_image_g_pca <- stats::prcomp(loaded_image_g, center=FALSE)
loaded_image_b_pca <- stats::prcomp(loaded_image_b, center=FALSE)
```

### Scree Plot {.tabset}

It is used to plots the order of the percentage of explained variance of the original data by each principal component.

#### Red Channel

```{r, echo=FALSE, warning=FALSE, message=FALSE}
fviz_eig(loaded_image_r_pca, main = "")
```

#### Green Channel

```{r, echo=FALSE, warning=FALSE, message=FALSE}
fviz_eig(loaded_image_g_pca, main = "")
```

#### Blue Channel

```{r, echo=FALSE, warning=FALSE, message=FALSE}
fviz_eig(loaded_image_b_pca, main = "")
```

## Image Compression

```{r, include=FALSE}
loaded_image_pca <- list(loaded_image_r_pca, loaded_image_g_pca, loaded_image_g_pca)
e_values_r <- get_eigenvalue(loaded_image_r_pca) %>% dplyr::mutate_if(is.numeric, round, 2)
e_values_g <- get_eigenvalue(loaded_image_g_pca) %>% dplyr::mutate_if(is.numeric, round, 2)
e_values_b <- get_eigenvalue(loaded_image_b_pca) %>% dplyr::mutate_if(is.numeric, round, 2)
```


The size of compressed image is decided by the number of principal components chosen for compression. 

**Less number of principal components results in lower size of image but involves more loss of information.**

Number of principal components can be chosen based on percentage of variance explained.

### 90% Explained Variance

```{r}
explained_variance <- 90.00

components <- list()
components[[1]] <- e_values_r %>% 
  dplyr::filter(cumulative.variance.percent < explained_variance) %>% 
  nrow()
components[[2]] <- e_values_g %>% 
  dplyr::filter(cumulative.variance.percent < explained_variance) %>% 
  nrow()
components[[3]] <- e_values_b %>% 
  dplyr::filter(cumulative.variance.percent < explained_variance) %>% 
  nrow()

compressed_image <- sapply(1:length(loaded_image_pca), function(channel_index) {
  channel_pca <- loaded_image_pca[[channel_index]]
  n_components <- components[[channel_index]]
  
  compressed_channel <- channel_pca$x[, 1:n_components] %*% t(channel_pca$rotation[, 1:n_components])
  compressed_channel[compressed_channel > 1] <- 1
  compressed_channel[compressed_channel < 0] <- 0
  return(compressed_channel)
}, simplify = 'array')

jpeg::writeJPEG(compressed_image, paste0("./tmp/compressed_", explained_variance, ".jpeg"))

plotImage(list(compressed_image))
```

```{r, echo=FALSE, results='asis'}
original_size <- file.info(file_path)$size
compressed_size <- file.info(paste0("./tmp/compressed_", explained_variance, ".jpeg"))$size
cat("* ", "Size of original image is ", original_size, " bytes.\n\n", sep='')
cat("* ", "Size of compressed image is ", compressed_size, " bytes.\n\n", sep='')
cat("* ", "Compression percentage: ", round(100 * (compressed_size - original_size) / original_size, 2), "%", sep='')
```


### 95% Explained Variance

```{r}
explained_variance <- 95.00

components <- list()
components[[1]] <- e_values_r %>% 
  dplyr::filter(cumulative.variance.percent < explained_variance) %>% 
  nrow()
components[[2]] <- e_values_g %>% 
  dplyr::filter(cumulative.variance.percent < explained_variance) %>% 
  nrow()
components[[3]] <- e_values_b %>% 
  dplyr::filter(cumulative.variance.percent < explained_variance) %>% 
  nrow()

compressed_image <- sapply(1:length(loaded_image_pca), function(channel_index) {
  channel_pca <- loaded_image_pca[[channel_index]]
  n_components <- components[[channel_index]]
  
  compressed_channel <- channel_pca$x[, 1:n_components] %*% t(channel_pca$rotation[, 1:n_components])
  compressed_channel[compressed_channel > 1] <- 1
  compressed_channel[compressed_channel < 0] <- 0
  return(compressed_channel)
}, simplify = 'array')

jpeg::writeJPEG(compressed_image, paste0("./tmp/compressed_", explained_variance, ".jpeg"))

plotImage(list(compressed_image))
```

```{r, echo=FALSE, results='asis'}
original_size <- file.info(file_path)$size
compressed_size <- file.info(paste0("./tmp/compressed_", explained_variance, ".jpeg"))$size
cat("* ", "Size of original image is ", original_size, " bytes.\n\n", sep='')
cat("* ", "Size of compressed image is ", compressed_size, " bytes.\n\n", sep='')
cat("* ", "Compression percentage: ", round(100 * (compressed_size - original_size) / original_size, 2), "%", sep='')
```

### 99% Explained Variance

```{r}
explained_variance <- 99.00

components <- list()
components[[1]] <- e_values_r %>% 
  dplyr::filter(cumulative.variance.percent < explained_variance) %>% 
  nrow()
components[[2]] <- e_values_g %>% 
  dplyr::filter(cumulative.variance.percent < explained_variance) %>% 
  nrow()
components[[3]] <- e_values_b %>% 
  dplyr::filter(cumulative.variance.percent < explained_variance) %>% 
  nrow()

compressed_image <- sapply(1:length(loaded_image_pca), function(channel_index) {
  channel_pca <- loaded_image_pca[[channel_index]]
  n_components <- components[[channel_index]]
  
  compressed_channel <- channel_pca$x[, 1:n_components] %*% t(channel_pca$rotation[, 1:n_components])
  compressed_channel[compressed_channel > 1] <- 1
  compressed_channel[compressed_channel < 0] <- 0
  return(compressed_channel)
}, simplify = 'array')

jpeg::writeJPEG(compressed_image, paste0("./tmp/compressed_", explained_variance, ".jpeg"))

plotImage(list(compressed_image))
```

```{r, echo=FALSE, results='asis'}
original_size <- file.info(file_path)$size
compressed_size <- file.info(paste0("./tmp/compressed_", explained_variance, ".jpeg"))$size
cat("* ", "Size of original image is ", original_size, " bytes.\n\n", sep='')
cat("* ", "Size of compressed image is ", compressed_size, " bytes.\n\n", sep='')
cat("* ", "Compression percentage: ", round(100 * (compressed_size - original_size) / original_size, 2), "%", sep='')
```

### 99.99% Explained Variance

```{r}
explained_variance <- 99.99

components <- list()
components[[1]] <- e_values_r %>% 
  dplyr::filter(cumulative.variance.percent < explained_variance) %>% 
  nrow()
components[[2]] <- e_values_g %>% 
  dplyr::filter(cumulative.variance.percent < explained_variance) %>% 
  nrow()
components[[3]] <- e_values_b %>% 
  dplyr::filter(cumulative.variance.percent < explained_variance) %>% 
  nrow()

compressed_image <- sapply(1:length(loaded_image_pca), function(channel_index) {
  channel_pca <- loaded_image_pca[[channel_index]]
  n_components <- components[[channel_index]]
  
  compressed_channel <- channel_pca$x[, 1:n_components] %*% t(channel_pca$rotation[, 1:n_components])
  compressed_channel[compressed_channel > 1] <- 1
  compressed_channel[compressed_channel < 0] <- 0
  return(compressed_channel)
}, simplify = 'array')

jpeg::writeJPEG(compressed_image, paste0("./tmp/compressed_", explained_variance, ".jpeg"))

plotImage(list(compressed_image))
```

```{r, echo=FALSE, results='asis'}
original_size <- file.info(file_path)$size
compressed_size <- file.info(paste0("./tmp/compressed_", explained_variance, ".jpeg"))$size
cat("* ", "Size of original image is ", original_size, " bytes.\n\n", sep='')
cat("* ", "Size of compressed image is ", compressed_size, " bytes.\n\n", sep='')
cat("* ", "Compression percentage: ", round(100 * (compressed_size - original_size) / original_size, 2), "%", sep='')
```

### 100% Explained Variance

```{r}
compressed_image <- sapply(1:length(loaded_image_pca), function(channel_index) {
  channel_pca <- loaded_image_pca[[channel_index]]
  n_components <- e_values_r %>% nrow()
  
  compressed_channel <- channel_pca$x[, 1:n_components] %*% t(channel_pca$rotation[, 1:n_components])
  compressed_channel[compressed_channel > 1] <- 1
  compressed_channel[compressed_channel < 0] <- 0
  return(compressed_channel)
}, simplify = 'array')

jpeg::writeJPEG(compressed_image, paste0("./tmp/compressed_", explained_variance, ".jpeg"))

plotImage(list(compressed_image))
```

```{r, echo=FALSE, results='asis'}
original_size <- file.info(file_path)$size
compressed_size <- file.info(paste0("./tmp/compressed_", explained_variance, ".jpeg"))$size
cat("* ", "Size of original image is ", original_size, " bytes.\n\n", sep='')
cat("* ", "Size of compressed image is ", compressed_size, " bytes.\n\n", sep='')
cat("* ", "Compression percentage: ", round(100 * (compressed_size - original_size) / original_size, 2), "%", sep='')
```

# References

1. https://ourarchive.otago.ac.nz/bitstream/handle/10523/7534/OUCS-2002-12.pdf?sequence=1&isAllowed=y

2. https://stattrek.com/matrix-algebra/covariance-matrix.aspx

3. https://en.wikipedia.org/wiki/Eigendecomposition_of_a_matrix

4. http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp/

5. https://rpubs.com/aaronsc32/image-compression-principal-component-analysis